---
title: 'Assignment 2: Instrumental Variables'
author: "Thammarat Bunjaroenphornsuk 22037148 & Iván Felipe Moreno Guzmán 19438013"
date: "2024-05"
output:
  pdf_document: default
  html_document:
    df_print: paged
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
library(haven)      # To use data sets from Stata and other programs
library(tidyverse)  # To manage data sets
library(stargazer)  # to produce well-formatted output tables
library(AER)        # to run IV regressions and obtain some data 
library(knitr)      # To knit 
options(scipen = 999)

```

# Part 1

*You want to assess the relationship between education and income (measured as consumption expenditure) in Malawi. The dataset iv dataset1 contains information on current household consumption expenditure and completed years of schooling for 70,000 individuals from 70 schools. These individuals are former students who were randomly selected from school records and contacted to receive information on their current living standards. Assume that all schools are subject to the same rules (children start school at age 6 and the highest possible number of completed school years is 12 in all schools) and that treatment effects are homogeneous. When interpreting your results, focus on causality, effect size, and statistical significance.*

```{r dataset1, results='asis'}
data <- haven::read_dta("iv_dataset1.dta")


## This first chunk of code just explore the data and the particular variables of interest
stargazer(as.data.frame(data), title="Summary",
          style = "aer")


p1 <- ggplot(data = data, aes(x = completed_school_years)) + 
    geom_histogram(bins = 12, fill = 'salmon', color = 'black') +
    ggtitle("")
p2 <- ggplot(data , aes(x = consexp)) + 
    geom_histogram(fill = '#2D9596', color = 'black') +
    ggtitle("")

ggpubr::ggarrange(p1,p2, ncol=2,
                  label.y = "Count" )

data_summary <- data.frame(
    Name = names(data),
    Class = sapply(data, function(x) paste(class(x),collapse = ",")),
    NAs = sapply(data, function(x) sum(is.na(x))),
    Unique_Values = sapply(data, function(x) length(unique(x))),
    stringsAsFactors = FALSE
  )

stargazer(as.data.frame(data_summary), title="Data Outlook",
          summary = F, style = "aer",
          column.sep.width = "2pt", 
          rownames = F)

```


## Question 1

*Run an OLS regression to investigate the effect of completed school years on consumption expenditure and interpret your results. In general, why is this approach unlikely to give you an unbiased estimate of the causal effect of educational attainment on consumption expenditure? What is an example of a likely potential source of endogeneity, and to what direction does this bias lead and why? Be as specific as possible.*

```{r firstOLS, results='asis'}
ols_reg <- lm(consexp~completed_school_years,data = data)

stargazer(ols_reg,  
          title="OLS Regression", type="latex", 
          df=FALSE)

```

*The result indicates that for each additional year of schooling, consumption expenditure is expected to increase by 31,747.580 units, holding other factors constant and it is statistically significant at 1%. (-> we can compare it with the original distribution of the variable, for example with the mean and the sd) 
The OLS regression approach is unlikely to give an unbiased estimate of the causal effect due to two potential endogeneity issues. Firstly, this model omits socioeconomic status of families, such as family income. Higher familiy income might provide better educational opportunities and also have higher consumption expenditures (indicating higher income) independently of education. The bias as a result of the omited variable would be positive. This means the OLS estimate of the effect of education on consumption expenditure would be inflated. Secondly, innaccurate reporting of years of schooling can bias the estimates. This may occur when individuals incorrectly remember the exact number of years they attended school or even might intentionally misreport their years of schooling to appear more educated. The estimated coefficient on schooling years would be biased toward zero, underestimating the true effect of education on consumption expenditure.*

## Question 2

*One of the key reasons for the low attendance and early school dropouts in Malawi are health problems. About 35% of schools were provided with a Water, Sanitation, and Hygiene (WASH) program, which has increased average student success in these schools. Based on this information, perform an instrumental variable (IV) regression to asymptotically recover the causal effect of completed school years on consumption expenditure and interpret your results.*

```{r IV, results='asis'}

TSLS <- ivreg(consexp~completed_school_years|WASHprogram, data = data)
stargazer(TSLS, type="text")
```
The IV estimate for the effect of completed school years on consumption expenditure is 21,356.170. This suggests that each additional year of schooling increases consumption expenditure by approximately 21,356.170 units with 1% level of significant. The use of the WASH program as an instrument allows us to interpret the estimated effect as causal, assuming the instrument is valid.

## Question 3

*Which treatment effect are you estimating in this setting and why?*

*Given the assumption of homogeneous treatment effects in the study setting, the instrumental variable regression estimates the Average Treatment Effect (ATE). This is because, under the homogeneity assumption, the effect of an additional year of schooling on consumption expenditure is constant across all individuals. Therefore, the causal effect estimated from the IV regression using the WASH program as an instrument reflects the average effect of education on consumption expenditure for the entire population.*

## Question 4

*Which assumptions are necessary for the IV approach of task 2? First state the assumptions in general terms and then apply them to the specific example of this assignment. Discuss the validity of the assumptions and, to the extent possible, perform a formal test. What bias do you expect in your IV estimates and why?*

To consider an instrumental variable as valid, it must fulfill two conditions: exclusion restriction, relevance and exogeneity. The first one implies that the instrument affects the dependent variable (consumption expenditure) only through the endogenous explanatory variable (education) and must not be correlated with the error term in the outcome equation. The WASH program improves education (years of schooling), which in turn affects consumption expenditure, but it should not directly affect consumption expenditure, or influences consumption expenditure only through its effect on completed school years. This assumption is generally not directly testable. We rely on contextual and theoretical justification. For example, if the WASH program impacts consumption expenditure only by increasing completed school years, the assumption holds. If the program also affects health directly and health status also determines performance to seek incomes, it might violate this assumption. In addition, if the WASH program was not randomly assigned, it might be allocated based on factors that also affect consumption expenditure. For example, schools in wealthier areas might be more likely to receive the WASH program, and these areas might also have higher consumption expenditures due to higher incomes. Therefore, the validity of the exclusion restriction is weakened by this case.

The second condition is that the instrument must be correlated with the endogenous explanatory variable (education in this case). Schools with the WASH program must have higher average completed school years compared to schools without the program. This condition is usually revised with the significance of the first stage, regressing the independent variable with the instrument and other controls. 

```{r IVtest}
summary(TSLS, diagnostic=T)
```
According to the test results, the F-statistic for the weak instruments test is 1,715.06, which is much greater than 10, indicating that the instrument (WASH program) is strongly correlated with the endogenous variable (completed school years). This confirms the relevance of the instrument. The Wu-Hausman test compares the IV estimates to OLS estimates to check for endogeneity. The test statistic is 28.45 with a p-value of 9.64e-08, which is highly significant. This indicates that endogeneity is a concern, justifying the use of IV over OLS. The Sargan test is typically used to test the overidentifying restrictions when there are multiple instruments. Since there is only one instrument in this model, the Sargan test is not applicable.

Given the high F-statistic in the weak instruments test, we expect little to no bias from weak instruments. The primary source of potential bias would come from a violation of the exclusion restriction, which we have to assume holds based on the context provided. If the WASH program affects consumption expenditure through channels other than education, it could bias the estimates. However, without evidence to the contrary, we proceed under the assumption that our IV estimates are unbiased.

## Question 5

*Now assume that the WASH program was randomly assigned. Does this change your answer with respect to the validity of the IV assumptions? If yes, how?*

*The random assignment of the WASH program significantly strengthens the validity of the IV assumptions, particularly the exclusion restriction assumption. This makes the IV estimates more reliable and credible, as the primary concern about selection bias (more prevalance of WASH program in richer area) that induces potential correlation between the instrument and unobserved factors, is addressed by randomization. The relevance of the instrument is already well-supported by the data. The random assignment of the WASH program should still result in a significant effect on completed school years, as seen from the strong F-statistic in the weak instruments test (1,715.06). Therefore, the relevance assumption remains satisfied and unchanged.*

## Question 6

*You know that sex and age of former students are important predictors of income, so you decide to control for them in your regressions. Repeat the regressions from task 1 and task 2 with these additional controls and interpret your results. Do your results change compared to before? Why?Why not?*

```{r,  results='asis'}
TSLS_with_controls <- ivreg(consexp~completed_school_years+sex + age|WASHprogram + sex + age,data=data)
OLS_with_controls <- lm(consexp~completed_school_years + sex + age, data = data)

# show results 
stargazer(ols_reg, TSLS, OLS_with_controls, TSLS_with_controls, type="text")
```
*According to the results, the coefficient for completed school years remains positive and significant across all models. The magnitude of the effect slightly increases when controlling for sex and age in both the OLS and IV regressions. This suggests that part of the effect of education on consumption expenditure is slightly mediated through the factors of sex and age. Thus, these controls help isolate the effect of education by holding constant other factors that influence consumption expenditure.

Moreover, The R square and Adjusted R square values increase when adding controls, indicating that the model with controls explains a larger proportion of the variance in consumption expenditure. Also, the standard errors slightly decreases slightly for the estimates in the model with controls, suggesting an improvement in the model's precision. This means that adding controls for sex and age provides a more precise estimate of the effect of education on consumption expenditure, and the robustness of the results across different specifications supports the validity of the findings.*

# Part 2

*You are evaluating a randomized controlled trial (RCT) in 250 primary schools in India that aims to tackle micronutrient deficiencies. As intervention, an NGO provided schools with a “Lucky Iron Leaf”, a small iron cooking tool that infuses your meals with iron to help prevent and treat iron deficiency. Your data is provided in the dataset iv dataset2. Your research goal is to investigate the effect of the iron ingot provision on school performance. For this purpose, of the 250 schools, 125 were randomly selected to receive “Lucky Iron Leaf”. At the end of the year, student performance was assessed using a standardized test. When looking at the data, you find out that not all schools in the treatment group actually used the iron ingot, and you suspect that treatment effects may be heterogeneous.*\

we cannot assumme homo - ITT

## Question 1

*Calculate the Intention-to-Treat effect (ITT) of iron ingot provision on student test scores and interpret your result. How does this effect differ from the conventional ATE? Why may the ITT still be of interest in the described example?*

lecture 5:
"One option is measure the ITT, i.e. the causal effect of treatment
assignment regardless of actual participation (does not require
further assumptions due to random assignment)"

evaluate program - not just giving b

```{r}
data2 <- haven::read_dta("iv_dataset2.dta")
```


Effect is on the ones who receive the treatment 

## Question 2

*Based on the estimated ITT, derive a consistent estimate for the causal effect of the treatment on compliers and interpret your result. For this, please do not use any IV regression function. Instead, calculate the effect using your ITT estimate with the Indirect Least Squares approach. Standard errors can be ignored for this task. Briefly comment on the results.*

```{r}
### VERSION A
# reduced form and first stage
FS_A <- lm(X~Z) # first stage
RF_A <- lm(Y~Z) # reduced form

beta_1_A <-  coef(RF_A)[2] / coef(FS_A)[2] # reduced form / first stage
```


## Question 3

*Replicate your results from task 2 using a “manual” Two-Stage-Least-Squares approach (2SLS), i.e. do not use any IV regression functions. Standard errors can be ignored for this task.*

lecture 5 - slide 7 

## Question 4

*Which assumptions must be fulfilled for your estimate from task 2 and task 3 to be a local average treatment effect? Under which assumption is LATE identical to ATE? Please be specific to the analysis at hand!*


